---
title: "lab_08"
author: "derek willis"
date: "2023-04-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
install.packages("ggrepel")
library(tidyverse)
library(lubridate)
library(tidycensus)
library(ggplot2)
library(ggrepel)
```



## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

```{r}
pgco_911_calls <- read_csv("data/prince_georges_2022_overdoses.csv") |>
  mutate(date = date(datetime),
         month = month(datetime, label = TRUE),
         week = week(datetime))  

pgco_911_calls 
```


## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below.

**A1.** 

The months with the highest number of 911-related overdose calls in Prince George's County do not align neatly with the months with the highest number of overdose calls in Baltimore County. 

In Baltimore County, overdose calls are highest in mid-summer; in Prince George's County, July brought the second-highest number of overdose calls while August brought the third-lowest. Conversely, Prince George's County saw its highest call total in December, while Baltimore County saw its second-lowest call total in December. 

```{r}
pgco_911_calls_by_month <- pgco_911_calls |>
  group_by(month) |>
  summarise(
    month_calls = n()) |>
  mutate(pct_of_yearly_total = month_calls/sum(month_calls)*100) |> 
  arrange(desc(pct_of_yearly_total))

  
pgco_911_calls_by_month
```

```{r}
pgco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,month_calls), weight=month_calls)) +
  theme_economist() +
  labs(
    title="PG County Overdose Calls Spike in July, Holiday Season",
    x = "month",
    y = "total calls",
    caption = "source: Prince George's County County"
    
  )
```
```{r}
pgco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=(month), weight=month_calls)) +
  theme_economist() +
  labs(
    title="PG County Overdose Calls Spike in July, Holiday Season",
    x = "month",
    y = "total calls",
    caption = "source: Prince George's County County"
    
  )
```


**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** 

The 28th week of the year -- which began on July 9, 2022 -- has the highest call total. This may have some relationship with summer holiday festivities. The 51st week of the year -- which began on December 17th, 2022 -- saw the second-highest overall total, which likely has to do with holiday-time malaise and/or holiday bonus checks; the preceding week saw the fifth-highest total. The 48th week of the year -- which began on November 26th -- saw the third-highest total, and the preceding week saw the fourth-highest total; those figures are likely connected to Thanksgiving. 

Every week in the top 10 for overdose calls is roughly aligned with a holiday (Christmas, Halloween, Thanksgiving, Fourth of July) and/or midsummer, with two exceptions: the 10th week (early March) and the 39th week (the end of September/beginning of October). Those could be attributed to unusually potent batches sold during those weeks, but it is difficult to determine. 

It is not uncommon for overdoses to spike on and around major holidays.

In this dataset, the 53rd week of the year is only a single day: December 31. It appears as an outlier in our graph because it does not reflect a seven-day week.

This data only reflects calls; it does not reflect deaths or total overdoses.


```{r}
pgco_911_calls_by_week <- pgco_911_calls |>
  group_by(week) |>
  summarise(
    week_calls = n()) |>
  mutate(pct_of_yearly_total = week_calls/sum(week_calls)*100) |> 
  arrange(desc(pct_of_yearly_total))

  
pgco_911_calls_by_week
```

```{r}
pgco_911_calls_by_week |>
  ggplot() +
  geom_line(aes(x=week, y=week_calls)) + 
  theme_economist() +
  scale_x_continuous(breaks=seq(0,53,by=5))+
  labs(
    title="Overdose Calls Spike Around Holidays",
    x = "week",
    y = "total calls",
    caption = "source: Prince George's County County")

```

**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  

Based on the chart, it appears that lower-income (relative to the PG County) zip codes see higher overdose-related 911 call volumes, with some notable outliers. One such outlier is zip code 20710, a largely Black and Hispanic zip code in Bladensburg with a median household income of roughly $55,000 and the lowest-income zip code in the dataset. Zip code 20710 was the source of 25 calls -- fewer than most other lower-income zip codes. 

I'm struck by the range of call volumes by zip code - from fewer than 10 to just over 100 -- and by the fact that the demographics of lower-income zip codes with the highest call totals are relatively similar to the demographics of zip codes with relatively low call totals. Is it possible to discern a reason for the differences between those zip codes?

I would attempt to identify call locations with greater specificity to determine whether there are any notable drivers of overdose calls in the zip codes with the highest call totals - a recovery home, a house frequented by drug users, a hospital, etc. I would then look for corollary locations (recovery homes, hospitals, etc.) in lower-income zip codes with lower call totals. I would also contact social services and harm reduction outreach providers to get more detailed profiles of both categories of zip codes.

Once I have a better sense of what to look for, I would find an outreach worker willing to take me to notable locations to interview drug users about their communities/community dynamics. 

```{r}
pgco_911_calls_by_zip <- pgco_911_calls |>
  group_by(zipcode) |>
  summarise(
    total_calls = n()) |>
  mutate(pct_of_total = total_calls/sum(total_calls)*100) |> 
  arrange(desc(pct_of_total))

pgco_911_calls_by_zip

```

```{r}
acs5 <- load_variables(2019, "acs5", cache = TRUE)
View(acs5)
```

```{r}
md_medhouseholdincome <- get_acs(geography = "zcta",
              variables = c(medhouseholdincome = "B19013_001"),
              state = "MD",
              year = 2019)


md_medhouseholdincome
```
```{r}
pg_911_calls_with_medhouseholdincome <- pgco_911_calls_by_zip |> left_join(md_medhouseholdincome, by=c('zipcode'='GEOID'))

pg_911_calls_with_medhouseholdincome
```


```{r}
p <- ggplot(pg_911_calls_with_medhouseholdincome, aes(x=total_calls, y=estimate)) + geom_point() +
  geom_point(shape = 19, color = 'blue', size=1) 

p +geom_text_repel(aes(label=zipcode), nudge_x = 0.1, nudge_y = 0.1, size = 2.5) +
  ggtitle("Higher Income Zip Codes see fewer overdose-related 911 calls") +
    labs(
    title="Lower-Income Zip Codes See Higher Overdose-Related 911 Call Volumes",
    x = "Overdose-related 911 calls in 2022",
    y = "Median Household Income",
    caption = "Source: Prince George's County"
  )
```

