---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
knitr::opts_chunk$set(echo = TRUE)
options(scipen=999)

# Load the tidyverse
install.packages('tidyverse')
library(tidyverse)

# Load janitor
install.packages('janitor')
library(janitor)
```
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

```{r}
md_agency_payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=373564)
```
# Load in the Maryland grants and loan data
```{r}
md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv", guess_max=17765)
```


## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1:

I found 81 records, including payments to both Prince George's County Public Schools and the Prince George's County Board of Education, which presumably can only spend that revenue on public education. If I do not include the variations of the Prince George's County Board of Education, I find 66 records.

I did not include the PG SCD, which I believe stands for Soil Conservation District.

```{r}
# cleaning function
cleaned_md_agency_payments <- md_agency_payments |>
  clean_names()

# display the cleaned dataset
cleaned_md_agency_payments
```

```{r}
pg_any <- cleaned_md_agency_payments |> filter((str_detect(vendor_name, "PG") | str_detect(vendor_name, "PRINCE GEORGE'S COUNTY") | str_detect(vendor_name, "PRINCE GEORGES") | str_detect(vendor_name, "PRINCE")))
nrow(pg_any)

pg_any

```


```{r}
school_any <- cleaned_md_agency_payments |> filter((str_detect(vendor_name, "PUBLIC SCHOOL") |str_detect(vendor_name, "SCHOOL DISTRICT") | str_detect(vendor_name, "PGPS")))

school_any
```

```{r}
boe_any <- cleaned_md_agency_payments |> filter((str_detect(vendor_name, "BOARD OF EDUCATION") | str_detect(vendor_name, "BD OF EDUCATION") | str_detect(vendor_name, "BOE")))

boe_any
```

```{r}
pg_schools_no_boe <- cleaned_md_agency_payments |> 
filter((str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOLS") | str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL")))

nrow(pg_schools_no_boe)

pg_schools_no_boe
```


```{r}
pg_schools <- cleaned_md_agency_payments |> filter((str_detect(vendor_name, "PRINCE GEORGES PUBLIC") |str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGE'S COUNTY") | str_detect(vendor_name, "PRINCE GEORGES COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL") | str_detect(vendor_name, "PRINCE GEORGES PUBLIC SCHOOLS") | str_detect(vendor_name, "BD OF EDUCATION OF PG CO") | str_detect(vendor_name, "BOARD OF EDUCATION OF PRINCE GEORGES COUNTY MD") | str_detect(vendor_name, "PRINCE GEORGES COUNTY SCHOOL")))

nrow(pg_schools)

pg_schools
```


Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:

Prince George's County Public Schools received the largest total amount in payments from state agencies during the most recent complete school year: the 2022 school year. 

The annual payments to Prince George's Public Schools grew at a relatively slow rate -- even declining in the wake of the 2008 recession -- until 2015, when payments suddenly surged. After another brief dip, the total annual payments to Prince George's County Public Schools rose rapidly during and after the COVID-19 pandemic. 

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(
    total_payments = sum(amount),
  ) |>
  arrange(desc(total_payments))
```

```{r}
pg_schools |>
  group_by(fiscal_year) |>
  summarise(
    total_payments = sum(amount),
  ) |>
  arrange(desc(fiscal_year))
```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.

As in the case of the Prince George's County School District, several vendors received payments under multiple variations of their name. 

Examples: 

1. WHITING TURNER CONTRACTING CO is almost certainly the same as WHITING TURNER CONTRACTING COMPANY. That said, WHITING TURNER CONTRACTING CO and WHITING TURNER CONTRACTING COMPANY appear twice with different zip codes, forming two pairs of CO/COMPANY with matching zip codes.

2. PLANO-CONDON, LLC is likely the same vendor as PLANO-COUDON LLC. This may be a case of mispelling, given that Plano-Coudon is a large DMV-area contractor and Plano-Condon is not. 

3. Based on a quick search of online business records, CANNON DESIGN INC and CANNON WASHINGTON INC appear to be the same entity. 

Once confirming that all zip codes in the original dataset are in a standardized format, cleaning the data to merge identical vendors would require comparing core fragments of a vendor name -- CANNON, for instance -- and their associated zip code. If the zip codes match, the two fragments likely refer to the same vendor and the records could be merged. 

In some cases -- like Whiting Turner Contracting Company -- two versions of a vendor may appear with different zip codes. In that case, I am inclined not to merge seemingly matching vendors with non-matching zip codes out of an abundance of caution, though the non-matching zip codes could represent a warehouse and a headquarters (for example).

At minimum, this could require the use of a case_when function, but the solution may be more complicated than I am currently able to write. 


```{r}
umd_any <- cleaned_md_agency_payments |> filter((str_detect(agency_name, "UMD") | str_detect(agency_name, "UNIVERSITY")))

nrow(umd_any)

umd_any

```

```{r}
umd_capital_projects <- cleaned_md_agency_payments |> filter((str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")))

nrow(umd_capital_projects)

umd_capital_projects
```

```{r}
umd_capital_projects |>
  group_by(vendor_name,vendor_zip) |>
  summarise(
    total_payments = sum(amount),
  ) |>
  arrange(desc(total_payments))
```


Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4:

Maryland funded far more stem cell research in 2009 than it did in 2022, and despite fluctuations, the average grant funding from 2009-2014 was lower than average grant funding from 2015-2022. 

```{r}
# cleaning function
cleaned_md_grants_loans <- md_grants_loans |>
  clean_names()

# display the cleaned dataset
cleaned_md_grants_loans
```

```{r}
stem_cell_grants <- cleaned_md_grants_loans |> 
filter((str_detect(description, "STEM CELL") | str_detect(description, "Stem Cell") | str_detect(grantee, "STEM CELL")))
nrow(stem_cell_grants)

stem_cell_grants
```

```{r}
stem_cell_grants |>
  group_by(fiscal_year) |>
  summarise(
    total_grants = sum(amount),
  ) |>
  arrange(desc(fiscal_year))
```
```


Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees there are, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: 

My code turns up 18 grantees, though nine of those are either the University of Maryland (the entire institution) or a department, office or program within the University of Maryland. To clean up these results, I would first need to decide whether to treat all departments, offices and programs of the University as the same grantee. If not, I could leave each of them as a stand-alone, but I would still need to use a case_when function -- and possibly other cleaning tools -- to standardize the variety of names for the entire university, including UNIVERSITY OF MD,COLLEGE PARK and UNIVERSITY OF MARYLAND COLLEGE PARK, as a single grantee to ensure I have only one record per grantee.

One grantee -- OVALTZ INC -- appears only once, but its name may be a typo. Maryland business records show an Ovalz, Inc. registered in College Park, but no Ovaltz, Inc. 

Arguably more interesting is Zymetis, Inc., a bioengineering company that has patented used for a marine organism (the Z-microbe) found in the Chesapeake Bay that can convert sygar, starch and cellulose into useable chemicals and fuels, ostensibly to support the "sustainable chemicals and fuels industry" (further research needed, because some 'sustainable fuels' aren't especially safe).

```{r}
```{r}
# cleaning function
cleaned_md_grants_loans <- cleaned_md_grants_loans |>
  clean_names() |> 
  mutate(zip5 = str_sub(zip_code, start=1L, end=5L)) |>
  mutate(zip5 = case_when(
    zip5 == "Vario" ~ NA,
    zip5 == "UB7 O" ~ NA,
    zip5 == "UB7  " ~ NA,
    .default = zip5
  ))

# display the cleaned dataset
cleaned_md_grants_loans
```

```{r}
umd_college_park_zip <- cleaned_md_grants_loans |> 
filter( zip5 == 20742)

nrow(umd_college_park_zip)

umd_college_park_zip

```
```{r}
umd_college_park_zip |>
  group_by(grantee) |>
  summarise(
    total_grants_loans = sum(amount),
  ) |>
  arrange(desc(total_grants_loans))
```

```

