---
title: "lab_07"
author: "derek willis"
date: "2023-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

-   A Census API key

## Load libraries and establish settings

You'll need to load two packages for this: the tidyverse and tidycensus.

**Task** load these two packages

```{r}
# Turn off scientific notation
options(scipen=999)
library(tidyverse)

# Load the tidyverse.
library(tidycensus)
```

## Setup Census API

You'll need your Census API Key:

**Task** set your API Key if you haven't already. You won't need to install or overwrite it.

```{r echo=FALSE}
```

## Load and view ACS variables

You'll also need to load the list of American Community Survey variables from the 2021 5-year estimates so you can identify the codes you need:

**Task** load the variables from the 2021 5-year ACS (which is called "acs5") and save them to a dataframe

```{r}
acs5 <- load_variables(2021, "acs5", cache = TRUE)
View(acs5)
```

## Answer questions

**Q1** What is the Maryland jurisdiction with the lowest median age, according to the ACS 2021 5-year estimates? You will need to find the variable for median age first. 
**A1** 

Baltimore City has the lowest estimated median age (35.7 years) among Maryland jurisdictions, followed closely by Wicomico County. 

```{r}
md_medage <- get_acs(geography = "county",
              variables = c(medage = "B01002_001"),
              state = "MD",
              year = 2021)

md_medage <- md_medage |> arrange(estimate) |> top_n(24, estimate)

md_medage

```

**Q2** Which Maryland counties have a median age of at least 50? You can use the dataframe you just made to answer this, but you must write code to display only those counties with a median age of at least 50. 
**A2** 

Only Talbot County and Worcester County have median ages of at least 50. Both are eastern shore retirement destinations. 

I think it's somewhat notable that the second-oldest and second-youngest counties are adjacent to one another. 

```{r}
md_medage |> 
  filter(estimate >= 50)|>
  arrange(desc(estimate)) 

```

**Q3** We're interested in knowing more about Montgomery County zip codes where overdose calls are coming from and the demographic characteristics they possess. In particular, we want to calculate a rate of calls per 1,000 population. To do this, you'll need to:

1. Load the Montgomery County 911 overdose calls data.
2. Create a new dataframe with the total number of calls for each zip code.
3. Find the ACS variable pertaining to the total 18 and up population (there are several choices here; you want the simplest one).
4. Get from the ACS data from all Maryland zip codes using that variable.
5. Join the ACS data to the Montgomery County zip code totals dataframe and rename the estimate column to a meaningful name you've given the ACS variable.
6. Add a column to the joined dataframe that calculates a rate of calls per 1,000 population.

Which zip code has the highest rate of calls? Which has the lowest?


**A3** 

Zip code 20877 in Gaithersburg has the highest number of calls per capita (only counting those 18 and older) at roughly 1.7 - a significantly lower number than the highest per capita number of calls in lower-income counties. While the 20877 zip code is less affluent than average for Montgomery County, it is still wealthier than the national average (when measured by household income). 

Zip code 20707 has the lowest number of calls per capita in the dataset, though most of the zip code is in Prince George's County. The next-lowest number of calls per capita come from zip code 20832 in Olney, which is wealthier than the Montgomery County average. 


```{r}
moco_2022_od <- read_csv("data/montgomery_2022_overdoses.csv")

moco_2022_od$zip <- as.character(moco_2022_od$zip)
```

```{r}
moco_2022_od_zip <- moco_2022_od |>
    group_by(zip) |>
    summarize(count = n()) |> 
    arrange(desc(count))

  
moco_2022_od_zip
```


```{r}
md_over_18 <- get_acs(geography = "zcta",
              variables = c(md_over_18 = "B09021_001"),
              state = "MD",
              year = 2019)

md_over_18
```
```{r}
moco_2022_od_18up_zip <- inner_join(moco_2022_od_zip, md_over_18, by=c("zip" = "GEOID")) |>
  rename(pop_over_18 = estimate)

```

```{r}
moco_2022_od_18up_zip <- moco_2022_od_18up_zip |>
  mutate(
    od_calls_pc = (count/pop_over_18)*1000
  )|>
  arrange(desc(od_calls_pc))

moco_2022_od_18up_zip

```


**Q4** Using [CensusReporter.org](https://censusreporter.org/), examine the zip codes with the highest and lowest rates and compare and contrast their demographic profiles. Is there a story here?

**A4** 

Since the per capita call total for zip code 20707 uses the population of the full zip code but the call total from the small portion of the zip code within Montgomery County (https://www.montgomerycountymd.gov/gis/Resources/Files/pdfs/zipcodes_esize.pdf), the figure isn't useful. 

Zip code 20832 in Olney isn't surprising given the area's estimated median household income, but it is noteworthy that Olney has a lower number of calls per capita than even wealthier zip codes like 20854 in Potomac or 20851 in Chevy Chase. 

The 20877 zip code in Gaithersburg has a substantially higher poverty rate (15.4%) than Montgomery County (7.9%). Its population is also substantially more Hispanic than average for Montgomery County (42% vs. 20%). Given that roughly 40 percent of adults in the county speak Spanish at home, it may be worthwhile to investigate whether Montgomery County's emergency response, harm reduction service providers and MOUD providers can adequately provide services in Spanish. That said, we don't have demographic information on people who experience overdoses in Montgomery County at hand, so we can't be certain that the demographics of callers in the 20877 zip code match the demographics of the zip code's population. 

I also think that zip code 20852 in Rockville and North Bethesda - the zip code with the second-highest number of calls per capita - stands out for having an estimated median household income far closer to (but still slightly below) the Montgomery County average and a lower poverty rate than the two other zip codes with the highest number of overdose-related 911 calls per capita. 


**Q5** Go back to your joined dataframe from Q3 and add two more columns representing the minimum and maximum populations based on the estimate and margin of error, then calculate per capita rates for each. Do you get the same two zip codes for the highest and lowest as you did before? Does that give you more or less confidence in your answers to Q3?

**A5** 

Even when adjusted for the margin of error, the data returns the same zip codes when calculating the highest and lowest number of OD 911 calls per capita. That gives me more confidence in my answers to Q3, though I remain unconvinced that zip code 20707 is a useful datapoint in this analysis. 

```{r}
moco_2022_od_18up_zip <- moco_2022_od_18up_zip |>
    mutate(
      pop_max = (pop_over_18 + moe),
      pop_min = (pop_over_18 - moe),
      od_calls_pc_max = (count/pop_max) * 1000,
      od_calls_pc_min = (count/pop_min) * 1000,
    )

moco_2022_od_18up_zip

```

```{r}
moco_2022_od_18up_zip |>
  arrange(desc(od_calls_pc_max))
```
```{r}
moco_2022_od_18up_zip |>
  arrange(desc(od_calls_pc_min))
```


